### 2018-07-25 chapters/01-introduction/intro.tex
Clear unused lines and use grammarly checks

Translations from generic \LaTeX{} to \gls*{cas} are difficult to realize since the semantic information are not explicitely given in the input. The \gls*{nist} has created a set of semantic \LaTeX{} macros~\cite{DLMF:Macros}. Each macro ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. These macros embeds necessary semantic information into \LaTeX{} expressions. One example of such a macro is given table~\ref{tab:JacobiP-usecase} for the semantic \LaTeX{} representation for the Jacobi polynomial. The macros provide isolated access to important parts of the mathematical function, such as the arguments.

>>>>

Translations from generic \LaTeX{} to \gls*{cas} are difficult to realize since the semantic information is not explicitly given in the input. The \gls*{nist} has created a set of semantic \LaTeX{} macros~\cite{DLMF:Macros}. Each macro ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. These macros embed necessary semantic information into \LaTeX{} expressions. One example of such a macro is given table~\ref{tab:JacobiP-usecase} for the semantic \LaTeX{} representation for the Jacobi polynomial. The macros provide isolated access to important parts of the mathematical function, such as the arguments.

### 2018-07-25 chapters/02-background/related-work.tex
Clear unused lines and use grammarly checks

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most of the \gls*{cas} provide simple functions to import and export mathematical \LaTeX{} expressions%
\footnote{The selected \gls*{cas} Maple, Mathematica, Matlab, and SageMath provide import and/or export functions for \LaTeX:\\
Maple, \url{http://www.maplesoft.com/support/help/Maple/view.aspx?path=latex} seen 06/2017;\\
Mathematica, \url{https://reference.wolfram.com/language/tutorial/GeneratingAndImportingTeX.html} seen 06/2017;\\
Matlab, \url{https://www.mathworks.com/help/symbolic/latex.html} seen 06/2017;\\
SageMath, \url{http://doc.sagemath.org/html/en/tutorial/latex.html} seen 06/2017}. %
Those tools have two essential problems. They are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. However, for more complex expressions, e.g., the Jacobi polynomial in \cref{tab:JacobiP-usecase}, the import functions fail. The second problem appears in the export tools. Mathematical expressions in \gls*{cas} are fully semantic, otherwise the \gls*{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information gets lost, because generic \LaTeX{} is not able to carry semantic information. In consequence of these two problems, an exported expression cannot be imported to the same system again in most cases (except those simple expressions described above). Our tool should solve these problems and provide round trip translations between \LaTeX{} and \gls*{cas}.

The semantics must be well known before an expression can be translated. There are two naiv approaches to solve that problem: (1) someone could specify the semantic information during the writing process (pre-defined semantics) or (2) the translator can find out the right semantic information in general mathematical expressions before it translates the expression. So called \textit{interactive documents}\footnote{There is no adequate definition what interactive documents are. However, this name is widely used to describe electronic document formats that allow interactivity to change the content in real time.}, such as the \gls*{cdf}~\parencite{CDF:Wolfram} by Wolfram Research or the \textit{worksheets} by \Maple{}, try to solve this problem with approach (2) and allow to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow to perform computations during \gls*{cas}. This gives the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible for the user. Similar to the \JOBAD{} architecture other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}. All of them demonstrate the potential for the education system.

%All of these systems carry semantic information. Most of them create their own way to make this possible. The web based documents mostly use standardized ways, such as the content \gls*{mathML}, a specialization of \gls*{mathML}~\parencite{MathML}, that focuses on the semantics of an expression rather than any particular rendering for the expression. Since \LaTeX{} is the de facto standard to write mathematical expressions in documents there are projects to extend \LaTeX{} in a way that enable to carry the semantic information directly in \LaTeX. There are two known projects that created a semantic version of \LaTeX: \sTeX{}~\parencite{sTeX} developed by Michael Kohlhase and the \Macro s developed by Bruce Miller. Our translator uses the \Macro s rather than \sTeX. A full explanation of this decision will be given in \cref{sec:semantics}.

%However, \textit{interactive documents}, web based or not, have an essential problem: all of them create an entire new document format. We want to create an independent lite translation tool for \LaTeX. Once we achieve this goal, an integration into \textit{interactive documents} might be possible, besides many other possible applications.

@@

%Obviously, the first example is not as powerful as a \gls*{cas} would be. However, \textit{sagetex} does not really solve our problem of the scientific workflow. The writer still needs to translate expressions manually. The only difference is that the author can write computations directly into the \TeX -file. However, these approaches have the potential for a powerful combination with our translation tool.

%In summary, today's translation tools are either specialized for one specific \gls*{cas} (such as the import and export functions of the \gls*{cas}) or they completely avoid the translation process by creating a new document format (such as \textit{interactive documents}). Our goal is to provide a lite translation tool for mathematical \LaTeX{} expressions and multiple \gls*{cas}.

There exists two approaches that allow to embed semantic information within \LaTeX{} expressions by using custom macros. Namely, \sTeX{}~\parencite{sTeX} developed by Kohlhase and the \Macro s developed by Miller~\parencite{DLMF:Macros}. This paper shows that it is possible to develop a context-free translation tool using the semantic macros introduced by these two projects. \sTeX{} aims to embed semantic information for general mathematics with a comprehensive set of macros. The macro set developed by Miller introduces new macros for special functions, orthogonal polynomials, and mathematical constants. Each of these macros ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. Therefore, we call these semantic macros \Macro s. These semantic macros are internally used in the \gls*{dlmf} and the \gls*{drmf}. Because of the linked definitions to the \gls*{dlmf} and the special macros for functions and mathematical constants this paper using the \Macro s for performing translations to \gls*{cas} instead of using \sTeX.

>>>>

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most of the \gls*{cas} provide simple functions to import and export mathematical \LaTeX{} expressions%
\footnote{The selected \gls*{cas} Maple, Mathematica, Matlab, and SageMath provide import and/or export functions for \LaTeX:\\
Maple, \url{http://www.maplesoft.com/support/help/Maple/view.aspx?path=latex} seen 06/2017;\\
Mathematica, \url{https://reference.wolfram.com/language/tutorial/GeneratingAndImportingTeX.html} seen 06/2017;\\
Matlab, \url{https://www.mathworks.com/help/symbolic/latex.html} seen 06/2017;\\
SageMath, \url{http://doc.sagemath.org/html/en/tutorial/latex.html} seen 06/2017}. %
Those tools have two essential problems. They are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. However, for more complex expressions, e.g., the Jacobi polynomial in \cref{tab:JacobiP-usecase}, the import functions fail. The second problem appears in the export tools. Mathematical expressions in \gls*{cas} are fully semantic. Otherwise the \gls*{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information gets lost, because generic \LaTeX{} is not able to carry semantic information. In consequence of these two problems, an exported expression cannot be imported to the same system again in most cases (except those simple expressions described above). Our tool should solve these problems and provide round-trip translations between \LaTeX{} and \gls*{cas}.

The semantics must be well known before an expression can be translated. There are two main approaches to solve that problem: (1) someone could specify the semantic information during the writing process (pre-defined semantics) or (2) the translator can find out the right semantic information in general mathematical expressions before it translates the expression. So-called \textit{interactive documents}\footnote{There is no adequate definition what interactive documents are. However, this name is widely used to describe electronic document formats that allow interactivity to change the content in real time.}, Such as the \gls*{cdf}~\parencite{CDF:Wolfram} by Wolfram Research or the \textit{worksheets} by \Maple{}, try to solve this problem with the approach (2) and allow to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web-based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow to perform computations during \gls*{cas}. This gives the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible to the user. Similar to the \JOBAD{} architecture other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}. All of them demonstrate the potential of the education system.

@@

There exist two approaches that allow embedding semantic information within \LaTeX{} expressions by using custom macros. Namely, \sTeX{}~\parencite{sTeX} developed by Kohlhase and the \Macro s developed by Miller~\parencite{DLMF:Macros}. This paper shows that it is possible to develop a context-free translation tool using the semantic macros introduced by these two projects. \sTeX{} aims to embed semantic information for general mathematics with a comprehensive set of macros. The macro set developed by Miller introduces new macros for special functions, orthogonal polynomials, and mathematical constants. Each of these macros ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. Therefore, we call these semantic macros \Macro s. These semantic macros are internally used in the \gls*{dlmf} and the \gls*{drmf}. Because of the linked definitions to the \gls*{dlmf} and the special macros for functions and mathematical constants this paper using the \Macro s for performing translations to \gls*{cas} instead of using \sTeX.

### 2018-07-25 chapters/03-translator/backward-trans.tex
Clear unused lines and use grammarly checks

%\subsection{Internal Data Structure}
%While our forward translation is realized for the \texttt{1D} input representation in \Maple, this representation is not suitable for a backward translation. Since \Maple{} uses it's own syntax and programming language, we would need a new parser for \Maple{} expressions to realize a backward translation tool. Of course, \Maple{} has this parser implemented. But it is not possible to use the parser without using \Maple{} itself. Therefore, an installed version of \Maple{} is mandatory for our backward translation.

%As already mentioned, \Maple{} has several ways to represent an expression. The \texttt{1D} and \texttt{2D} representations are used as input representations. Besides those, there are internal data structures to perform all kinds of computations on the expression. Internally, each expression is handled and stored as a \gls*{dag}. The \Maple{} \gls*{dag} is very similar to an expression tree. But for efficiency, it does not store copies of an object. Consider the integral from (\ref{eq:maple-input})
%\begin{equation}\label{eq:maple-input}
%\int\displaylimits_0^\infty \frac{\pi+\sin(2x)}{x^2} dx.
%\end{equation}

%As already mentioned, the \Maple{} \gls*{dag} has no copies of objects. Therefore, the '$x$' in \eqref{eq:maple-input-2} only appears once in the \gls*{dag}.
%\begin{figure}[ht]
%\centering
%\includegraphics[scale=0.7]{DAG.pdf}
%\caption{The \Maple{} DAG of equation~\eqref{eq:maple-input}}
%\label{fig:maple-dag}
%\end{figure}

%The \Maple{} \gls*{dag} can be visualized in a graph, such as in figure~\ref{fig:maple-dag}. But the internal representation looks a bit different.

@@

%\Maple{} has also another representation, similar to the \gls*{dag}, that allows multiple copies of the same object in its representation. This representation is called \inertF{} and is more similar to an expression tree than \Maple's \gls*{dag}. The \inertF{} represents the \gls*{dag} as a tree by splitting object copies and displays the tree in a list structure. Each node (or element in the list) has the prefix \texttt{\_Inert\_XXX}, that indicates the node is in the \inertF. The suffix \texttt{XXX} is the type of the node. Some of the important types are specified in table~\ref{tab:maple-types}. Note that this is only a subset of all possible types.

@@

%The \texttt{OpenMaple} \gls*{api} is implemented for different programming languages such as \texttt{C}, \texttt{Java} and \texttt{Visual Basic}. Since our forward translation and the \gls*{mlp} is implemented in \texttt{Java}, we use the \texttt{Java} \gls*{api} for the backward translation. The \texttt{Java} \gls*{api} has some limitations to interact with \Maple's \gls*{dag}.

%For the backward translation, we use the \inertF. This representation can be converted to a nested list version and the \texttt{OpenMaple} Java \gls*{api} can handle lists. Therefore, the backward translation firstly converts the \inertF{} to a nested list structure before any translation process starts.

%The following figure~\ref{fig:inertform-list} shows the \inertF{} and the nested list of the polynomial $x^2+x$.

%\begin{figure}[ht]
%\centering
%\begin{minipage}{.45\linewidth}
%\centering
%\begin{tabular}{lll}
%\multicolumn{3}{l}{\_Inert\_SUM(}\\
%& \multicolumn{2}{l}{\_Inert\_POWER(}\\
%& & \_Inert\_NAME("x"),\\
%& & \_Inert\_INTPOS(2)\\
%& ),\\ 
%& \multicolumn{2}{l}{\_Inert\_NAME("x")}\\
%)
%\end{tabular}
%\end{minipage}
%\begin{minipage}{.45\linewidth}
%\centering
%\begin{tabular}{lll}
%\multicolumn{3}{l}{[\_Inert\_SUM,}\\
%& \multicolumn{2}{l}{[\_Inert\_POWER, }\\
%& & [\_Inert\_NAME, "x"],\\
%& & [\_Inert\_INTPOS, 2]\\
%& ],\\ 
%& \multicolumn{2}{l}{[\_Inert\_NAME, "x"]}\\
%]
%\end{tabular}
%\end{minipage}
%\caption{\inertF{} and the nested list representation of $x^2+x$.}
%\label{fig:inertform-list}
%\end{figure}

%In the nested list, the first element specifies the type of the current list while the following arguments store the arguments and optional attributes.

\subsection{Automatic Changes of Inputs in Maple}\label{subsec:maple-probs}
%Note that the backward translations usually do not have problems with ambiguous expressions. Every \gls*{cas} needs to solve ambiguities in its own representation, otherwise it could not compute the expression. Our backward translation from \Maple{} back to semantic \LaTeX{} is therefore based on already disambiguated expressions.

%The general walkthrough of our backward translation starts with a string of a \texttt{1D} \Maple{} input expression. As described above, this string is entered in \Maple{} to parse the expression and work on the nested list version of the \inertF{} afterwards. This causes some problems. Mostly because \Maple{} automatically evaluates input expressions immediately - even before we have a chance to take a look at the internal data structure of the input. For example, an input such as $\sin\left(\frac{\cpi}{2}\right)$ is immediately evaluated to $1$ and the internal data structure just shows us the positive integer value $1$ instead of the trigonometric function.

%Furthermore, \Maple{} also changes input expressions slightly, mostly because of missing data types for a specific expression. For example, the internal data structure has no type to represent fractions except the \texttt{RATIONAL} type, which only allows integer values for the numerator and denominator. Thereby, an expression like $\frac{x}{2}$ is automatically changed to $x\frac{1}{2}$. There is no way no avoid such simple arithmetic changes in the expression, but we can avoid the automatic evaluations.

%Firstly, we summarize all known changes and evaluations \Maple{} automatically performs on input expressions.

\Maple{} evaluates inputs automatically and changes the input into an internal representation. This internal representation might look a bit different to the input. One example has already been given with figure~\ref{fig:internal-maple-dag}, where each summand of a sum is stored with a factor. Here is a list of all internal changes that occur for inputs.

@@

Since we want to keep a translated expression similar to the input expression, we implemented some cosmetic rules for the backward translations that solves or reduce the effects from the list of changes above. 
\begin{itemize}
\item We use unevaluation quotes to supress evaluations of the input.
\item We perform reordering of factors and summands so that negative factors appears in front of the summand. This gives us the opportunity to translate $x - y$ to $x - y$ instead of $x + y \cdot (-1)$.
\item We introduced the new internal data types \texttt{MYFLOAT} and \texttt{DIVIDE} to translate floats and fractions in more convenient notations.
\end{itemize}

%Since we want to keep a translated expression similar to the input expression, we use unevaluation quotes during the whole translation process. Unevaluation quotes also solve the problems with square roots and $n$-th roots. In \Maple's \texttt{1D} input representation, a square root is a function call (\texttt{sqrt(x)}) and the unevaluation quotes prevent evaluations on functions. Therefore, a backward translation simply needs to translate a square root as a normal function. Hence, we do not have problems with the internal representation of roots in \Maple.

%Because of the absent data type for subtractions, a long sum with negative terms could be difficult to read. We change the order of constants (such as an integer value) to be in front of products. Therefore, an expression such as '$-y$' is not represented by '$y\cdot (-1)$', but by '$(-1) \cdot y$'. The translator now needs to check if there is a leading '$-1$' and can translate it to '$-y$' rather than to '$(-1) \cdot y$'.

%Floating point numbers in scientific notation also causes a problem. Consider the input expression $0.41$ as the nested list \inertF{} representation
%\begin{equation}
%\texttt{[\_Inert\_FLOAT, [\_Inert\_INTPOS, 41], [\_Inert\_INTNEG, 2]]}.
%\end{equation}

%The backward translator is implemented in \texttt{Java}. To translate such an expression back to $0.41$, computations become necessary. Of course, it makes sense to perform those computations in the \gls*{cas} (in that case, in \Maple) rather than in the translation program. Therefore, we implemented a procedure\footnote{\textit{Procedures} in \Maple{} are small programs similar to methods and functions in \gls*{oop} languages.} to automatically convert \texttt{FLOAT} nodes to a string representation of the floating point number. We created a new type, called \texttt{MYFLOAT} to represent such numbers. With this approach, we avoided computations in the translation process.

%For fractions, we use a similar approach and introduce a new internal data type \texttt{DIVIDE}. The implementation of \texttt{DIVIDE} was difficult. On the one hand, some expressions with negative exponent should be displayed as a fraction while others should should be displayed with the negative exponent. We follow the same rule as \Maple{} follows internally to display fractions. When the exponent is a negative numeric element, the expression should be displayed as a fraction. If the exponent is something else, we uses the negative exponent representation. Table~\ref{tab:maple-fracs} shows two examples with negative exponents. One of them is displayed as a fraction (exponent '$-2$') and the other is not (exponent '$-2y$'). We implement the same rule for our newly developed \texttt{DIVIDE} data type. Note that with the new created data type, we are still not able to avoid changes from $\frac{x}{2}$ to $x\frac{1}{2}$, but we are able to work with more general fraction expressions.

%\begin{table}[ht]
%\centering
%\begin{tabular}{p{2.6cm}cc}
%	\hline
%	& Negative numeric exponent & General negative exponent\\
%	\hline
%	\texttt{1D} \Maple{} input expression & $\texttt{(x-1)\^{}(-2)}$ & \rule{0pt}{0.9\normalbaselineskip}$\texttt{(x-1)\^{}(-2*y)}$\\
%	How \Maple{} displays the input & $\displaystyle\frac{1}{(x-1)^2}$ & $\displaystyle(x-1)^{-2y}$\\
%	\hline
%\end{tabular}
%\caption{Different styles to display negative exponents depending on the type of the exponent. \Maple{} displays expressions with negative numeric exponents as fractions, while other negative exponents are not displayed as fractions.}
%\label{tab:maple-fracs}
%\end{table}

%We implement all of these changes before the actual translation process starts. This helps keeping the translated expression similar to the input expression.

%\subsection{Implementation Details}\label{subsec:maple-impl}
%\begin{wrapfigure}{l}{0.485\textwidth}
%	\vspace{-12pt}
%	\centering
%	\includegraphics[clip, trim=0.1cm 0.1cm 0.1cm 0.1cm, scale=0.39]{BackwardTranslator.pdf}
%	\caption{A scheme of the backward translator and its specialized subtranslators.}
%	\label{fig:BackwardDel}
%	\vspace{-13pt}
%\end{wrapfigure}

%The general translation process runs through different states. Each translation process starts with the string representation of a \texttt{1D} \Maple{} input. This string will be converted to the internal nested list representation of our custom \inertF{} (see the previous subsection~\ref{subsec:maple-probs}). Since the \inertF{} is similar to an expression tree, it can be translated as the \gls*{mlp-pt} in the forward translation. Hence, we use the same approach of multiple subtranslators for the backward translation process.

%Figure~\ref{fig:BackwardDel} shows the subtranslators of the backward translation process. The structure is the same as for the forward translation, seen in figure~\ref{fig:ForwardDel}. The \verb|MapleInterface| represents the entry point of a translation process. Before the translation starts, the \Maple{} kernel needs to be initialized and prepared for the translation process by loading all custom procedures. Once everything is loaded, an expression is delegated to specialized subtranslators depending on the internal data type, defined by custom procedures and \Maple's \inertF.

%The following figure~\ref{fig:backward-trans} illustrates the backward translation process for the Jacobi polynomial example $P_n^{(\alpha , \beta)}(\cos(a\Theta))$ from table~\ref{tab:JacobiP-usecase}. The input expression is converted to the nested list representation of the \texttt{MyInertForm} (the customized \inertF{} of \Maple). Afterwards, each node of the tree is translated separately by specialized subtranslators (visualized by blue arrows). The translation of functions (bold blue arrows) is again realized by translation patterns, which define the correct position for each argument. The last step is used to put all translated arguments to the right position (visualized by red arrows).

The translation process than follows the same principle as for the forward translations. Since the syntax tree of \Maple{} is an expression tree, we do not need to implement special reordering or grouping algorithms to perform backward translations. Translations for functions are also realized via patterns and placeholders. Figure~\ref{fig:backward-trans} illustrates the backward translation process for the Jacobi polynomial example from table~\ref{tab:JacobiP-usecase}.

>>>>

@@

@@

\subsection{Automatic Changes of Inputs in Maple}\label{subsec:maple-probs}
\Maple{} evaluates inputs automatically and changes the input into an internal representation. This internal representation might look a bit different to the input. One example has already been given with figure~\ref{fig:internal-maple-dag}, where each summand of a sum is stored with a factor. Here is a list of all internal changes that occur for inputs.

@@

Since we want to keep a translated expression similar to the input expression, we implemented some cosmetic rules for the backward translations that solve or reduce the effects from the list of changes above. 
\begin{itemize}
\item We use unevaluation quotes to suppress evaluations of the input.
\item We perform reordering of factors and summands so that negative factors appear in front of the summand. This gives us the opportunity to translate $x - y$ to $x - y$ instead of $x + y \cdot (-1)$.
\item We introduced the new internal data types \texttt{MYFLOAT} and \texttt{DIVIDE} to translate floats and fractions in more convenient notations.
\end{itemize}

The translation process then follows the same principle as for the forward translations. Since the syntax tree of \Maple{} is an expression tree, we do not need to implement special reordering or grouping algorithms to perform backward translations. Translations for functions are also realized via patterns and placeholders. Figure~\ref{fig:backward-trans} illustrates the backward translation process for the Jacobi polynomial example from table~\ref{tab:JacobiP-usecase}.
